---
title: "Results Replication"
author: "Neel Roy"
date: "2023-06-22"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Here is my code:**

## Reformatting/Cleaning Up Data

### Raw Counts File

```{r}

cnts <- read.csv('raw_counts_Cooper_data_final.csv')

rownames(cnts) <- cnts$X

cnts$X <- NULL

cnts <- as.matrix(cnts)
```

This takes raw_counts file and corrects the columns and index of the dataframe, then converts the file from a data frame to a matrix.

### Meta Data

```{r}

column_data <- read.csv('Cooper_Data_Meta.csv')

rownames(column_data) <- column_data$Sample

column_data <- column_data[,c("Tissue","Treatment","Stage")]


#column_data$Tissue <- factor(column_data$Tissue)
#column_data$Treatment <- factor(column_data$Treatment)
#column_data$Stage <- factor(column_data$Stage)

#column_data$Treatment <- NULL
#column_data$Stage <- NULL
#column_data$Sample <- NULL

column_data
```

This takes the metadata, or column data, file and corrects the columns and index of the dataframe, then converts the file from a data frame to a matrix.

### Check if rownames and columnnames Match

```{r}

all(rownames(column_data) %in% colnames(cnts))

all(rownames(column_data) == colnames(cnts))
```

This checks if the rownames and column names match.

### DeSeq2

```{r}

library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cnts,
                              colData = column_data,
                              design = ~ Tissue)
dds
```

### **Pre-filtering**

```{r}

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

Prefiltering can improve visualization as features with no information are not plotted.

### Factor Levels

```{r}

dds$Tissue <- relevel(dds$Tissue, ref = "Benign_Tissue")
```

"By default, R will choose a *reference level* for factors based on alphabetical order. Then, if you never tell the DESeq2 functions which level you want to compare against (e.g. which level represents the control group), the comparisons will be based on the alphabetical order of the levels. There are two solutions: you can either explicitly tell *results* which comparison to make using the `contrast` argument (this will be shown later), or you can explicitly set the factors levels. In order to see the change of reference levels reflected in the results names, you need to either run `DESeq` or `nbinomWaldTest`/`nbinomLRT` after the re-leveling operation."

Setting Benign Tissue as the reference level.

## **Differential expression analysis**

```{r}

dds <- DESeq(dds)
resultsNames(dds)

#summary(res)
```

```{r}

res_Serous_EOC <- results(dds, name="Tissue_Serous_EOC_vs_Benign_Tissue")

res_Serous_EOC

resLFC_Serous_EOC <- lfcShrink(dds, coef="Tissue_Serous_EOC_vs_Benign_Tissue", type="apeglm")

resLFC_Serous_EOC
```

```{r}

res_Ascites <- results(dds, name="Tissue_Ascites_vs_Benign_Tissue")

res_Ascites

resLFC_Ascites <- lfcShrink(dds, coef="Tissue_Ascites_vs_Benign_Tissue", type="apeglm")

resLFC_Ascites
```

Using lfcShrik to make vizualization and ranking of genes better.

```{r}

resLFCOrdered_Serous_EOC <- resLFC_Serous_EOC[order(resLFC_Serous_EOC$pvalue),]

resLFCOrdered_Serous_EOC

resLFCOrdered_Ascites <- resLFC_Ascites[order(resLFC_Ascites$pvalue),]

resLFCOrdered_Ascites
```

Ordered by p-value.

```{r}

summary(resLFCOrdered_Serous_EOC)

summary(resLFCOrdered_Ascites)
```

### MA Plots

```{r}

plotMA(resLFCOrdered_Serous_EOC, ylim=c(-2,2))

idx_LFCOrdered_Serous_EOC <- identify(resLFCOrdered_Serous_EOC$baseMean, resLFCOrdered_Serous_EOC$log2FoldChange)
rownames(resLFCOrdered_Serous_EOC)[idx_LFCOrdered_Serous_EOC]

plotMA(resLFCOrdered_Ascites, ylim=c(-2,2))
```

```{r}

plotMA(resLFCOrdered_Serous_EOC, ylim=c(-2,2))

idx_LFCOrdered_Serous_EOC <- identify(resLFCOrdered_Serous_EOC$baseMean, resLFCOrdered_Serous_EOC$log2FoldChange)
rownames(resLFCOrdered_Serous_EOC)[idx_LFCOrdered_Serous_EOC]
```

```{r}

resNorm <- lfcShrink(dds, coef=3, type="normal")
resAsh <- lfcShrink(dds, coef=3, type="ashr")
resLFC <- lfcShrink(dds, coef=3, type="apeglm")

par(mfrow=c(1,3), mar=c(4,4,2,1))
xlim <- c(1,1e5); ylim <- c(-3,3)
plotMA(resLFC, xlim=xlim, ylim=ylim, main="apeglm")
plotMA(resNorm, xlim=xlim, ylim=ylim, main="normal")
plotMA(resAsh, xlim=xlim, ylim=ylim, main="ashr")
```

### **Plot Counts**

```{r}

plotCounts(dds, gene=which.min(resLFC_Serous_EOC$padj), intgroup="Tissue")

plotCounts(dds, gene=which.min(resLFC_Ascites$padj), intgroup="Tissue")

plotCounts(dds, gene=which.max(resLFC_Serous_EOC$padj), intgroup="Tissue")

plotCounts(dds, gene=which.max(resLFC_Ascites$padj), intgroup="Tissue")
```

### Data Transformations + Variance

```{r}

vsd <- vst(dds, blind=FALSE)
#rld <- rlog(dds, blind=FALSE)
ntd <- normTransform(dds)
library("vsn")
head(assay(vsd), 10)
meanSdPlot(assay(ntd))
meanSdPlot(assay(vsd))
#meanSdPlot(assay(rld))
```

### Heatmaps

```{r}

library("pheatmap")
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("Tissue","Treatment")])
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
#pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=FALSE,
#         cluster_cols=FALSE, annotation_col=df)
```

```{r}

sampleDists <- dist(t(assay(vsd)))

library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, vsd$type, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

### PCA

```{r}

plotPCA(vsd, intgroup=c("Tissue"))
```

### Random

```{r}

results(dds, contrast=c("Tissue","Serous_EOC","Benign_Tissue"))

```

#### More LFC Shrinkage

```{r}

resApeT <- lfcShrink(dds, coef=2, type="apeglm", lfcThreshold=1)
plotMA(resApeT, ylim=c(-3,3), cex=.8)
abline(h=c(-1,1), col="dodgerblue", lwd=2)
```

```{r}

resAshT <- lfcShrink(dds, coef=2, type="ashr", lfcThreshold=1)
plotMA(resAshT, ylim=c(-3,3), cex=.8)
abline(h=c(-1,1), col="dodgerblue", lwd=2)
```

#### Boxplot

```{r}

par(mar=c(8,5,2,2))
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2)
```

#### Dispersion Plot

```{r}

plotDispEsts(dds)
```

#### Other

```{r}

metadata(resLFC_Serous_EOC)$alpha

metadata(resLFC_Serous_EOC)$filterThreshold

plot(metadata(resLFC_Serous_EOC)$filterNumRej, 
     type="b", ylab="number of rejections",
     xlab="quantiles of filter")
lines(metadata(resLFC_Serous_EOC)$lo.fit, col="red")
abline(v=metadata(resLFC_Serous_EOC)$filterTheta)

resNoFilt <- results(dds, independentFiltering=FALSE)
addmargins(table(filtering=(resLFC_Serous_EOC$padj < .1),
                 noFiltering=(resNoFilt$padj < .1)))

```

```{r}

metadata(resLFC_Ascites)$alpha

metadata(resLFC_Ascites)$filterThreshold

plot(metadata(resLFC_Ascites)$filterNumRej, 
     type="b", ylab="number of rejections",
     xlab="quantiles of filter")
lines(metadata(resLFC_Ascites)$lo.fit, col="red")
abline(v=metadata(resLFC_Ascites)$filterTheta)

resNoFilt <- results(dds, independentFiltering=FALSE)
addmargins(table(filtering=(resLFC_Ascites$padj < .1),
                 noFiltering=(resNoFilt$padj < .1)))
```

```{r}

par(mfrow=c(2,2),mar=c(2,2,1,1))
ylim <- c(-2.5,2.5)
resGA <- results(dds, lfcThreshold=.5, altHypothesis="greaterAbs")
resLA <- results(dds, lfcThreshold=.5, altHypothesis="lessAbs")
resG <- results(dds, lfcThreshold=.5, altHypothesis="greater")
resL <- results(dds, lfcThreshold=.5, altHypothesis="less")
drawLines <- function() abline(h=c(-.5,.5),col="dodgerblue",lwd=2)
plotMA(resGA, ylim=ylim); drawLines()
plotMA(resLA, ylim=ylim); drawLines()
plotMA(resG, ylim=ylim); drawLines()
plotMA(resL, ylim=ylim); drawLines()
```

```{r}

#mcols(dds,use.names=TRUE)[1:4,1:4]

#substr(names(mcols(dds)),1,10) 

#mcols(mcols(dds), use.names=TRUE)[1:4,]

#head(assays(dds)[["mu"]])

#head(assays(dds)[["cooks"]])

#head(dispersions(dds))

#head(mcols(dds)$dispersion)

#sizeFactors(dds)

#head(coef(dds))

#attr(dds, "betaPriorVar")

#priorInfo(resLFC)
```

```{r}

library("AnnotationDbi")
library("org.Hs.eg.db")

final_data_Serous <- read.csv('Final_Data_LFC_Serous_EOC.csv')

final_data_Serous$symbol <- mapIds(org.Hs.eg.db, keys = final_data_Serous$X, column = 'SYMBOL', keytype = 'ENSEMBL')

final_data_Serous <- final_data_Serous[order(final_data_Serous$pvalue),]

head(final_data_Serous, 50)




final_data_Ascites <- read.csv('Final_Data_LFC_Ascites.csv')

final_data_Ascites$symbol <- mapIds(org.Hs.eg.db, keys = final_data_Ascites$X, column = 'SYMBOL', keytype = 'ENSEMBL')

final_data_Ascites <- final_data_Ascites[order(final_data_Ascites$pvalue),]

head(final_data_Ascites, 50)

```
